Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-08-10 ltx/data_paper/DataPaper.tex
edits to introduction

The human brain brain is arguably the most complicated system known to man. The development of a complete theory for its function is one of the greatest challenges faced by scientists today. To address this challenge, researchers have naturally used both theoretical and experimental frameworks to develop and test hypotheses investigating brain function. Experimental frameworks are typically designed to uncover causal relationships between the systems properties or parameters and its function (reverse engineering the brain). Theoretical studies are typically used in one of two ways. The first is to explain data acquired in an experiment and the second is to predict system behaviour.

@@

Although the predictive explanatory power of the theoretical models is rather vast when they are tuned to mimic experimental conditions, a limitation has been establishing a rigorous method for choosing model parameters in more general situations. For example, over the last decade it has come to light that there is a significant variability in parameters of neural models across subjects (ref Marder 2006). For example, it has been shown that a target network behaviour may be achieved by continuum of parameter combinations (Tail wagging the ref). We expect an analogous situation in disease states and pathological activity such as epileptic seizures. Therefore, for models to be clinically useful they must be subject-specific.

By assimilating experimental or clinical measurements with mathematical models one can infer unmeasured or latent system properties or parameters. Inference of patient-specific parameters has the the potential to revolutionise the treatment of disease. Typically, one can measure the blood-oxygen level dependent signal through fMRI or the electromagnetic fields of the brain through EEG or MEG. In clinical neurology, these measurement modalities give a kind or a `where and when' indication of the presence of a pathology, but provide little information on the causative mechanisms for disease. A classic example of this is in epilepsy management, where the disease is diagnosed using EEG. If electrical seizures are recorded, a medication plan is prescribed. The choice of medication is not based on the data acquired from the EEG, but based on the clinician's experience with similar cases and sometimes trial and error because similar electrographic seizure can have fundamentally different mechanisms of initiation (ref JTs paper). Ideally in epilepsy monitoring we would like to image the concentration of ions, the synaptic dynamics, the connectivity strength and structure etc. to better inform a treatment plan by understand physiological changes that lead to seizures. This is also highly desirable for epileptic seizure prediction..

@@

Another recent approach uses graph theory \cite{Albert2002,Bang-Jensen2010} for large-scale neural networks and employs realistic anatomical connectivity from the CoCoMac database \cite{Kotter2004} to model the brain's resting state activity \cite{Ghosh2008,Deco2009}. Each node in the network is described by a physiologically reasonable neural population model of their internal dynamics. The authors perform stability analysis on the network based on total connectivity and time delays in the presence of noise. Such an approach has the potential to model a virtual brain which can incorporate data from individual patients, giving insights into how patients' specific features can affect its large-scale neural dynamics \cite{Jirsa2010,Deco2010}.

>>>>

The human brain brain is arguably the most complicated system known to man. The development of a complete theory for its function is one of the greatest challenges faced by scientists today. To address this challenge, researchers have naturally used both theoretical and experimental frameworks to develop and test hypotheses. Experimental frameworks are typically designed to uncover causal relationships between the systems properties or parameters and its function (reverse engineering the brain). Theoretical studies are typically used in one of two ways. The first is to explain data acquired in an experiment and the second is to predict system behaviour.

@@

Although the predictive explanatory power of the theoretical models is rather vast when they are tuned to mimic experimental conditions, a major limitation has been in establishing a rigorous method for choosing model parameters in more general situations. For example, over the last decade it has come to light that there is significant variability in neural systems across subjects despite seemingly similar network activity (ref Marder 2006). It has been shown that a target network behaviour may be achieved by continuum of parameter combinations (Tail wagging the ref). We expect an analogous situation in disease states and pathological activity such as epileptic seizures. Therefore, for models to be clinically useful they must be subject-specific.

By assimilating experimental or clinical measurements with mathematical models one can infer unmeasured or latent system properties or parameters. Inference of patient-specific parameters has the the potential to revolutionise the treatment of disease. Typically, one can measure the blood-oxygen level dependent signal through fMRI or the electromagnetic fields of the brain through EEG or MEG. In clinical neurology, these measurement modalities give a kind or a `where and when' indication of the presence of a pathology, but provide little information on the causative mechanisms for disease. A classic example of this is in epilepsy management, where the disease is diagnosed using EEG by identifying electrographic seizures. If electrical seizures are recorded with EEG, a medication plan is prescribed. The choice of medication is not based on the data acquired with the EEG, but based on the clinician's experience and the patient’s circumstance, then the treatment plan is often modified based on trial and error. This process of searching for the best drug combination is often required similar electrographic seizures can have fundamentally different mechanisms of initiation (ref JTs paper). Ideally in epilepsy monitoring we would like to image the concentration of ions, the synaptic dynamics, the connectivity strength and structure etc. to better inform a treatment plan by understand physiological changes that lead to seizures. This is also highly desirable for epileptic seizure prediction.

@@

Another recent approach for large scale neural networks based on graph theory \cite{Albert2002,Bang-Jensen2010}, employs a realistic anatomical connectivity from CoCoMac data base \cite{Kotter2004} to model resting state activity in the brain \cite{Ghosh2008,Deco2009} where each node in the network is modelled by physiologically reasonable neural population model of their internal dynamics. They perform a stability analysis of the network in terms of total connectivity and time delays in the presence of noise. Such approach has a potential to model a virtual brain which can incorporate data from individual patient, giving insights into how patient specific features can affect its large scale neural dynamics \cite{Jirsa2010,Deco2010}.
